---
title: "Introduction to ACEBounds"
description: >
  How to use this package to find bounds.
output: html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to ACEBounds}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: "references.bib"
---

## Simulate Data 

To use for illustrative purposes, we simulate some data. The DAG corresponding to the model simulated from is presented below. $X,Y,U$ are all binary, while $Z1$ is categorical with three levels. 

```{r message = FALSE, warning = FALSE}
set.seed(184506)

library(ACEBounds)
library(magrittr)
library(dplyr)

sim_dat <- simulate_data(sample_size = 500,
                         IVs_ps = list(c(1, 1, 1)/3),
                         indIVs_on_X = 0.7)

library(ggdag)
plot_DAG(sim_dat) + scale_x_reverse()
```

The first 10 rows of the data simulated:

```{r}
knitr::kable(sim_dat$simulated_data, n = 10)
```


## With Individual Level Data

We create three data sets to illustrate the scenarios where either a one-sample or a two-sample study design is used.

```{r}
XYZ_data <- sim_dat$simulated_data

XZ_data <- sim_dat$simulated_data %>% select(X,Z1)
YZ_data <- sim_dat$simulated_data %>% select(Y,Z1)
```

### Trivariate data source

If we have observations of $(X,Y,Z)$, we can find the bounds on ACE as follows:

```{r}
probs <- probs_from_data(XYZ_data, X = X, Y = Y, Z = Z1)

get_bounds(zetas = probs$zetas)
```


If for whatever reason you want to get $P(X = 1 | Z = z)$ and $P(Y = 1 | Z = z)$ from a trivariate data source, that is also possible.

```{r}
bivariate_probs <- probs_from_data(XYZ_data, X = X, Y = Y, Z = Z1, data_format = "bivariate")

get_bounds(thetas = bivariate_probs$thetas,
           gammas = bivariate_probs$gammas)
```


### Bivariate data sources

If we have a two-sample study design, i.e. one data set contains observations on $(X, Z)$, and the other observations on $(Y,Z)$, we can still find bounds on the ACE.

```{r}
X_probs <- probs_from_data(XZ_data, X = X, Z = Z1)
Y_probs <- probs_from_data(YZ_data, Y = Y, Z = Z1)

get_bounds(thetas = X_probs$thetas,
           gammas = Y_probs$gammas)
```


## Assuming Monotonicity

If you feel comfortable assuming that $P(X = 1 | Z = z, U) \leq P(X = 1 | Z = z+1, U)$ for all values of $z$, we simply specify `x_mono = TRUE`:

```{r}
get_bounds(zetas = probs$zetas,
           x_mono = TRUE)

get_bounds(gammas = Y_probs$gammas,
           thetas = X_probs$thetas,
           x_mono = TRUE)
```

## With Summary Data Only

If you do not have individual level data available, you need to either provide vectors of $P(X = 1 | Z = z)$ and $P(Y = 1 | Z = z)$, or an array of $P(X = x, Y = y | Z = z)$. 

In the trivariate case, we take as an example the Vitamin A supplementation data from @balke_bounds_1997.

```{r results="asis"}
counts <- array(data = c(74, 0, 11514, 0,
                         34, 12, 2385, 9663),
                dim = c(2,2,2),
                dimnames = list(x = 0:1,
                                y = 0:1, 
                                z = 0:1))

pander::pandoc.table(counts, style = "rmarkdown")
```

We convert the counts to the respective proportions.

```{r results="asis"}
zetas <- counts / rep(apply(counts, 3, sum), each = 4)

pander::pandoc.table(zetas, style = "rmarkdown")
```

These can then be passed to the `zetas` argument of the `get_bounds()` function:

```{r}
get_bounds(zetas = zetas)
```

If we only had the values of $P(X = 1 | Z = z)$ and $P(Y = 1 | Z = z)$, these would be passed to the `thetas` and `gammas` arguments of the `get_bounds()` function, respectively:

```{r}
thetas <- c(0, 0.8)
gammas <- c(0.994, 0.996)


get_bounds(gammas = gammas,
           thetas = thetas)
```

