---
title: "Bounds using Two-Sample Data"
description: "Obtaining bounds on the ATE using two-sample data."
output: rmarkdown::html_vignette
params:
  n_cores: 2
vignette: >
  %\VignetteIndexEntry{Bounds using Two-Sample Data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

Here we will see how bounds on the ATE behaves when derived from a single IV in two-sample data. We know that the width of the bounds is bounded above by $2\cdot (1 - ST_X)$, when monotonicity of the effect of $Z$ on both $X$ and $Y$ holds. Here, we will see through simulations, that the same seems to hold without these assumptions. 

## Generate Distributions

We will consider IVs that take three levels. To generate values of $P(X = 1 | Z = z)$ we simply choose three values from a uniform distribution. The matrices obtained from the polymake program give us some constraints on the values of $P(Y = 1 | Z = z)$. These are included below. 

```{r warning = FALSE, message = FALSE}
library(tidyverse, quietly = TRUE)

ATEBounds:::matrices_from_polymake %>% 
  filter(!x_monotone, !y_monotone,
         data_format == "bivariate", 
         n_z_levels == 3) %>% 
  pull(matrix) %>% .[[1]] %>% 
  filter(alpha == 0) %>%
  select(where(~sum(abs(.x)) > 0)) %>%
  filter(rowSums(abs(.)) > 1) %>% 
  pander::pander()
```


The function `simulate_gammas_from_thetas` takes a vector of length 3 of values of $P(X = 1 | Z = z)$, and returns a vector of length 3 of values of $P(Y = 1 | Z = z)$ such that no constraint is violated. The comments in the function below refer to the corresponding row in the matrix of verifiable constraints above. 

```{r}
simulate_gammas_from_thetas <- function(thetas){

  gammas <- vector("numeric", length = 3)

  gammas[1] <- runif(1)

  gammas[2] <- runif(1,
                     min = max(0,
                               gammas[1] - thetas[1] - thetas[2], # row 3
                               gammas[1] + thetas[1] + thetas[2] - 2), # row 5
                     max = min(1,
                               gammas[1] - thetas[1] - thetas[2] + 2, # row 11
                               gammas[1] + thetas[1] + thetas[2]) # row 16
                     )

  gammas[3] <- runif(1,
                     min = max(0,
                               gammas[1] - thetas[1] - thetas[3], # row 4
                               gammas[1] + thetas[1] + thetas[3] - 2, # row 6
                               gammas[2] - thetas[2] - thetas[3], # row 12
                               gammas[2] + thetas[2] + thetas[3] - 2), # row 9
                     max = min(1,
                               gammas[2] - thetas[2] - thetas[3] + 2, # row 10
                               gammas[2] + thetas[2] + thetas[3], # row 13
                               gammas[1] - thetas[1] - thetas[3] + 2, # row 17
                               gammas[1] + thetas[1] + thetas[3])) # row 15
  return(gammas)
}
```

We generate $10,000$ sets of distributions.

```{r}
library(tidyverse)
library(ATEBounds)


set.seed(7226637)
sim_probs <- tibble(j = 1:10000) %>%
  mutate(thetas = map(j, ~runif(n = 3, min = 0, max = 1)), 
         gammas = map(thetas, simulate_gammas_from_thetas))
```


## Get Strength and Bounds.

Here, we define the strength of $Z$ on $X$ as $\max_{z_1,z_2} \left | P(X = 1 | Z = z_1) - P(X = 1 | Z = z_2) \right |$, and the strength of $Z$ on $Y$ similarly as $\max_{z_1,z_2} \left | P(Y = 1 | Z = z_1) - P(Y = 1 | Z = z_2) \right |$. 

We then calculate the strength of $Z$ on $X$, $Z$ on $Y$, and find the bounds. This is where we notice that some of the probability distributions result in bounds with an upper bound that is less than the lower bound. We generate extra random distributions such that the total number of distributions with correctly ordered bounds is $10,000$. 

```{r}
if(!file.exists(here::here("vignettes_data/bivariate_bounds.Rds"))){
  library(furrr)
  
  plan(multisession, workers = n_cores)
  
  bounds <- sim_probs %>% 
    mutate(strength_x = future_map_dbl(thetas, ~max(abs(outer(.x, .x, `-`)))),
           strength_y = future_map_dbl(gammas, ~max(abs(outer(.x, .x, `-`)))),
           bounds = future_map2(thetas, gammas,
                                ~get_bounds(thetas = .x,
                                            gammas = .y,
                                            stop = FALSE, 
                                            warning = FALSE)),
           violations = future_map_lgl(bounds, "constraints_violated"),
           intervals = future_map(bounds, "interval")) %>% 
    unnest_wider(intervals) %>% 
    mutate(width = upper - lower)
  
  n_extras <- 0
  
  while(sum(bounds$width >= 0) < nrow(sim_probs) & n_extras < 100){
    n_extras <- n_extras + 1
    
    plan(multisession, workers = 2)
    
    bounds_extra <- tibble(j = nrow(bounds) + 1:(nrow(sim_probs) - sum(bounds$width >= 0))) %>% 
      mutate(thetas = future_map(j, ~runif(n = 3, min = 0, max = 1), .options = furrr_options(seed = TRUE)), 
             gammas = future_map(thetas, simulate_gammas_from_thetas, .options = furrr_options(seed = TRUE)),
             strength_x = future_map_dbl(thetas, ~max(abs(outer(.x, .x, `-`)))),
             strength_y = future_map_dbl(gammas, ~max(abs(outer(.x, .x, `-`)))),
             bounds = future_map2(thetas, gammas,
                                  ~get_bounds(thetas = .x,
                                              gammas = .y,
                                              stop = FALSE, 
                                              warning = FALSE)),
             violations = future_map_lgl(bounds, "constraints_violated"),
             intervals = future_map(bounds, "interval")) %>% 
      unnest_wider(intervals) %>% 
      mutate(width = upper - lower)
    
    bounds <- bind_rows(bounds, bounds_extra)
  }
  
  write_rds(bounds, here::here("vignettes_data/bivariate_bounds.Rds"))
} else {
  bounds <- read_rds(here::here("vignettes_data/bivariate_bounds.Rds"))
}
```

The plot below shows the width plotted against the strength of $Z$ on both $X$ and $Y$. The black line has slope $-2$ and intercept $2$ -- i.e. it represents the upper bound on the width derived under the two monotonicity assumptions. 

```{r}
width_vs_strength <- bounds %>% 
  filter(!violations, upper > lower) %>% 
  ggplot(aes(x = strength_x, y = width)) + 
    geom_point(alpha = 0.2) + 
    geom_abline(slope = -2, intercept = 2) +
    lims(
      y = c(0,2)
    ) +
    labs(x = "Strength of IV",
         y = "Width of bounds") + 
    theme_bw()

width_vs_strength +
  labs(title = "Width of Two-Sample Bounds as function of strength")

ggsave(plot = width_vs_strength,
       filename = here::here("figures", "bivariate_width_vs_strength.tiff"), 
       dpi = 300, width = 4, height = 2.86)
```

Here are all the bounds:

```{r}
all_bounds_plot <- bounds %>% 
  filter(!violations, upper > lower) %>% 
  select(contains("strength"), upper, lower) %>% 
  arrange((upper + lower)/2) %>% 
  mutate(id = row_number()) %>% 
  ggplot(aes(x = id, y = (upper + lower)/2)) + 
    geom_errorbar(
      aes(ymin = lower, ymax = upper),
      width = 0.001, alpha = 0.05
    ) + 
    geom_line(
      aes(color = "Center of bounds"),
      alpha = 0.75
    )+ 
    labs(
      y = "ATE",
      x = "Simulation run",
      color = ""
    ) +
    guides(
      color = guide_legend(label.position = "top")
    ) +
    scale_color_manual(values = "red") +
    theme_bw() +
    theme(
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank(),
      legend.position = c(1, 0),
      legend.justification = c(1, 0),
      legend.background = element_blank(),
      legend.key = element_blank()
    )

all_bounds_plot +
  labs(title = "Bounds from ≈10,000 randomly generated distributions",
       caption = paste("Note:", sum(bounds$upper <= bounds$lower), "of the", nrow(bounds), "simulated distributions resulted in upper limits smaller than the lower limits. These have been removed."))

ggsave(plot = all_bounds_plot,
       filename = here::here("figures", "all_bivariate_bounds.tiff"),
       dpi = 300, width = 4, height = 2.86)
```

Many of these randomly generated distributions result in rather wide bounds, as can be seen from the table below.

```{r}
proportion_of_widths_greater_than <- bounds %>% 
  filter(width >= 0) %>% 
  group_by(Strength = cut(strength_x, breaks = c(0, 0.05, 0.1, 0.25, 0.5, 1), 
                          include.lowest = TRUE)) %>% 
  summarize(
    w = c(1, 0.75, 0.5),
    prop = map_dbl(c(1, 0.75, 0.5), ~mean(width > .x))
  ) %>% 
  mutate(w = glue::glue("w = {w}")) %>% 
  pivot_wider(names_from = w, values_from = prop)
  
pander::pander(proportion_of_widths_greater_than)
```

We save a few summaries for the manuscript.

```{r}
bounds %>% 
  summarize(
    n = n(),
    good = sum(upper >= lower),
    bad = sum(upper < lower)
  ) %>% 
  write_csv(here::here("tables/upper_less_than_lower.csv"))

write_csv(proportion_of_widths_greater_than,
          file = here::here("tables", "proportion_of_biv_widths_greater_than.csv"))
```

## Relate to real world data

We reproduce the plot of the width vs. strength, but include bounds from three real world examples. The first chunk below essentially summarizes the process laid out in `vignette("example_analysis")` and wraps it in a simple function. 

```{r message = FALSE, warning = FALSE}
library(TwoSampleMR)
library(magrittr)

find_intercept <- function(beta1, p_outcome, pz, interval = c(-5, 5), zs = 0:2){
  uniroot(f = function(beta0) sum((1+exp(-beta0-beta1*zs))^(-1) * pz) - p_outcome,
          interval = interval)
}

ao <- available_outcomes()

real_bounds <- function(exposure_id, outcome_id, ao = TwoSampleMR::available_outcomes()){
  
  exposure_experiment <- ao %>% filter(id == exposure_id)
  exposure_instruments <- extract_instruments(outcomes = exposure_experiment$id)
  
  outcome_experiment <- ao %>% filter(id == outcome_id)
  outcome_data <- extract_outcome_data(snps = exposure_instruments$SNP, outcomes = outcome_experiment$id)
  
  harmonized_data <- harmonise_data(exposure_instruments, outcome_data) # %>% as_tibble()
  
  pop_probs <- tibble(regression = c("outcome", "exposure"),
                      p_outcome = c(outcome_experiment  %$% { ncase / (ncase + ncontrol) },
                                    exposure_experiment %$% { ncase / (ncase + ncontrol) }))
  
  prep_for_bounds <- harmonized_data %>%
    rowwise() %>%
    mutate(samplesize.exposure = exposure_experiment$ncase + exposure_experiment$ncontrol,
           samplesize.outcome = outcome_experiment$ncase + outcome_experiment$ncontrol,
           ave_eaf = weighted.mean(x = c_across(c(eaf.exposure, eaf.outcome)),
                                   w = c_across(c(samplesize.exposure, samplesize.outcome)),
                                   na.rm = TRUE)) %>%
    select(SNP, contains("beta"), ave_eaf) %>%
    pivot_longer(contains("beta")) %>%
    separate(name, into = c("variable", "regression"), sep = "\\.") %>%
    pivot_wider(names_from = variable, values_from = value) %>%
    mutate(`P(Z = 2)` = (1 - ave_eaf)^2,
           `P(Z = 1)` = 2*ave_eaf*(1 - ave_eaf),
           `P(Z = 0)` = ave_eaf^2)
  
  include_intercept <- prep_for_bounds %>%
    left_join(pop_probs) %>%
    rowwise() %>%
    mutate(find_intercept = list(find_intercept(beta1 = beta,
                                                p_outcome = p_outcome,
                                                pz = c(`P(Z = 0)`, `P(Z = 1)`, `P(Z = 2)`),
                                                interval = c(-10, 10))),
           beta0 = find_intercept$root,
           `thetas/gammas` = list(1/(1+exp(-beta0 - 0:2*beta))))
  
  bivariate_bounds <- include_intercept %>%
    select(-find_intercept, -beta0, -p_outcome, -ave_eaf, coefs = beta) %>%
    mutate(regression = if_else(regression == "outcome", "gammas", "thetas")) %>% 
    pivot_wider(names_from = regression, values_from = c(`thetas/gammas`, coefs)) %>% 
    rename_with(.cols = starts_with("thetas"), 
                ~str_remove_all(.x, "thetas/gammas_")) %>% 
    rowwise() %>%
    mutate(bivariate_bounds = list(get_bounds(thetas = thetas, gammas = gammas,
                                              stop = FALSE, warning = FALSE)),
           interval = list(bivariate_bounds$interval),
           constraints_violated = bivariate_bounds$constraints_violated) %>%
    ungroup() %>%
    unnest_wider(interval)
  
  return(bivariate_bounds)
}
```

We find bounds of the average treatment effect of high cholesterol on incidence of heart attack, smoking on incidence of lung cancer, and obesity on incidence of heart attack. 

```{r fig.width = 4, fig.height = 4, dpi = 300}
real_data_biv_bounds <- tibble(exposure = c("Cholesterol", "Smoking", "Obesity"), 
                               outcome = c("Heart Attack", "Lung Cancer", "Heart Attack"),
                               exposure_id = c("ukb-a-108", "ukb-d-20116_0", "ieu-a-90"),
                               outcome_id = c("ukb-a-434", "ukb-d-40001_C349", "ukb-a-434")) %>% 
  rowwise() %>% 
  mutate(source = exposure,
         bounds = list(real_bounds(exposure_id = exposure_id, outcome_id = outcome_id, ao = ao))) %>% 
  ungroup() %>% 
  unnest(bounds) %>% 
  mutate(width = upper - lower,
         strength_x = map_dbl(thetas, ~max(abs(outer(.x, .x, "-")))))
```

With these in hand, we create a plot of the width of the bounds for all instruments against the strength of the instruments. 

```{r}
customvermillion <- rgb(213/255,94/255,0/255)
custombluegreen <- rgb(0/255,158/255,115/255)
customreddishpurple<-rgb(204/255,121/255,167/255) 
customblue <- rgb(0/255,114/255,178/255)

## Create plot-within-plot
width_vs_strength2 <- bounds %>% 
  filter(!violations, upper > lower) %>% 
  mutate(source = "Simulated") %>% 
  bind_rows(real_data_biv_bounds) %>% 
  ggplot(aes(x = strength_x, y = width, color = source)) + 
    geom_point(aes(alpha = source == "Simulated")) + 
    geom_abline(slope = -2, intercept = 2) +
    labs(x = "Strength of IV",
         y = "Width of bounds",
         color = "") + 
    scale_color_manual(values = c("Cholesterol" = customvermillion, 
                                  "Smoking" = custombluegreen,
                                  "Obesity" = customreddishpurple,
                                  "Simulated" = "black")) +
    scale_alpha_manual(values = c("TRUE" = 0.2, "FALSE" = 0.75)) +
    guides(color = guide_legend(override.aes = list(alpha = 1), nrow = 2),
           alpha = "none") +
    theme_bw() +
    theme(legend.position = "top")


zoomed_in <- width_vs_strength2 +
  coord_fixed(ratio = 0.25,
              xlim = c(0, 0.1), 
              ylim = range(real_data_biv_bounds$width)) + 
  scale_x_continuous(breaks = c(0, 0.05, 0.1),
                     expand = expansion(mult = 0, add = 0.001)) +
  labs(x = "", y = "") +
  guides(color = "none") +
  theme_bw() +
  theme(plot.background = element_blank(),
        plot.margin = unit(c(0.5,0.5, -0.1, -0.1), "lines"),
        axis.text.x = element_text(angle = 45, hjust = 1))


## 
box_x1 <- -0.01
box_x2 <- 0.1
box_y1 <- 0.94
box_y2 <- 1.275

zoomed_in_x1 <- 0.455
zoomed_in_x2 <- 1.0375
zoomed_in_y1 <- 1.11
zoomed_in_y2 <- 2.075

(pip_figure <- width_vs_strength2 + 
  coord_fixed(ratio = 0.5) +
  scale_y_continuous(limits = c(0, 2.075),
                     expand = expansion(mult = 0, add = c(0.05, 0.01))) +
  geom_rect(inherit.aes = FALSE,
            xmin = zoomed_in_x1, xmax = zoomed_in_x2,
            ymin = zoomed_in_y1, ymax = zoomed_in_y2,
            fill = "white") +
  annotation_custom(grob = ggplotGrob(zoomed_in), 
                    xmin = 0.43, 
                    ymin = 1.05) +
  geom_path(data = data.frame(x = c(box_x2, box_x1, box_x1, box_x2, box_x2, zoomed_in_x1), 
                              y = c(box_y1, box_y1, box_y2, box_y2, box_y1, zoomed_in_y1)), 
              aes(x = x, y = y),
              color = "red") + 
  geom_path(data = data.frame(x = c(box_x2, zoomed_in_x1),
                              y = c(box_y2, zoomed_in_y2)),
            aes(x = x, y = y),
            color = "red") +
  geom_rect(inherit.aes = FALSE,
            xmin = zoomed_in_x1, xmax = zoomed_in_x2, 
            ymin = zoomed_in_y1, ymax = zoomed_in_y2,
            fill = NA, color = "red")
)

write_rds(pip_figure, 
          file = here::here("vignettes_data/pip_figure.Rds"))

ggsave(filename = here::here("figures/bivariate_width_vs_strength_pip.tiff"),
       plot = pip_figure, 
       dpi = 300, width = 4, height = 4.7)
```

```{r eval = FALSE, include = FALSE}
bounds_w_bpbounds <- bounds %>% 
  filter(lower > upper) %>% 
  rowwise() %>% 
  mutate(bpbounds = list(tidy_bpbounds(bpbounds::bpbounds(p = as.table(rbind(1-gammas, gammas)), 
                                                          t = as.table(rbind(1-thetas, thetas)), 
                                                          fmt = "bivariate")) %>% 
                           filter(Assumption == "IV Inequality", `Causal parameter` == "ATE") %>% 
                           select(assumption_holds = `Assumption Holds`, bplower = `Lower bound`, bpupper = `Upper bound`)))

bounds_w_bpbounds %>% 
  ungroup() %>% 
  unnest(bpbounds) %>% 
  select(j, violations, bp_violations = assumption_holds, contains("lower"), contains("upper")) %>% 
  mutate(bp_violations = !bp_violations) %>% 
  filter(abs(upper - bpupper) > 10e-5) 

sim_zetas <- function(nz){
  zetas <- array(runif(4*nz), dim = c(2,2,nz), dimnames = list(x = 0:1, y = 0:1, z = 0:(nz-1)))
  
  return(zetas / rep(apply(zetas, 3, sum), each = 4))
}

sim_trivariate_probs <- tibble(j = 1:10000) %>%
  mutate(zetas = map(j, ~sim_zetas(nz = 3)))

trivariate_bounds <- sim_trivariate_probs %>% 
  rowwise() %>% 
  mutate(bounds = list(get_bounds(zetas = zetas, stop = FALSE, warning = FALSE)),
         violations = bounds$constraints_violated,
         intervals = list(bounds$interval))

trivariate_bounds %>% 
  ungroup() %>% 
  unnest_wider(intervals) %>% 
  filter(upper < lower) %>% 
  count(violations)
```

