---
title: "How to use the `get_bounds` function"
output: html_vignette
description: "On the usage of the `get_bounds` function to get bounds in various settings."
vignette: >
  %\VignetteIndexEntry{`get_bounds`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: "references.bib"
---


To get bounds on $\alpha$, one simply provides values of $\theta$ and $\gamma$, or $\zeta$ to the function `get_bounds()`. Let's generate some random values of $\theta$ and $\gamma$ for a three-leveled IV:

```{r}
set.seed(1406)
(thetas <- runif(3))
(gammas <- runif(3))
```

Here, `thetas` and `gammas` must be vectors of length 3 (since the IV is three-leveled). The i'th element of `thetas` must be $P(X = 1 | Z = i-1)$, and the i'th element of `gammas` must be $P(Y = 1 | Z = i-1)$. We can then feed these values to the `get_bounds()` function, and it will return a list with some results, including the interval given by the lower and upper bounds for $\alpha$. 

```{r}
library(ACEBounds)

get_bounds(thetas = thetas,
           gammas = gammas)
```


Similarly, you can choose to provide a 3 dimensional array of values to the `zetas` argument. This should have $P(Y = y, X = x | Z = z)$ as element $(y+1,x+1,z+1)$ (when $y \in \{0,1\}, x \in \{0,1\}, z \in \{0,...,(k-1)})$. Here is an example for an IV with four levels:

```{r}
probs <- c(.83, .05, .11, .01, 
           .88, .06, .05, .01, 
           .72, .05, .10, .13,
           .85, .02, .07, .06)

zetas <- array(probs, dim = c(2, 2, 4),
               dimnames = list(x = c(0, 1),
                               y = c(0, 1),
                               z = c(0, 1, 2, 3)))
zetas

get_bounds(zetas = zetas)
```


If the values you provide violate some of the constraints, an error is thrown by default. Here, $P(X = 1, Y = 0 | Z = 1) + P(X = 1, Y = 1 | Z = 2) = 1.08 > 1$, which violates the Balke and Pearl IV inequalities [@balke_bounds_1997]). 

```{r}
invalid_probs <- c(.83, .05, .11, .01, 
                   .88, .06, .05, .01, 
                   .72, .05, .20, .03)
invalid_zetas <- array(invalid_probs, dim = c(2, 2, 3),
                       dimnames = list(x = c(0, 1),
                                       y = c(0, 1),
                                       z = c(0, 1, 2)))
invalid_zetas
```

```{r eval = FALSE}
get_bounds(zetas = invalid_zetas)
```

```{r echo = FALSE, results="hide"}
get_bounds(zetas = invalid_zetas, stop = FALSE)
```


If you use this function programmatically, for example for a simulation study, you might want to change this behaviour. This can be achieved by setting `stop = FALSE`. A warning will be printed instead, but your program won't stop. The element `constraints_violated` in the output indicates that some constraints were not satisfied. 

```{r}
get_bounds(zetas = invalid_zetas, stop = FALSE)
```

This warning can also be disabled:

```{r}
get_bounds(zetas = invalid_zetas, stop = FALSE, warning = FALSE)
```

## References
