---
title: '"Power" Simulations'
description: ""
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{"Power" Simulations}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

Here we briefly show how the monte carlo integration that provides the data used in `vignette("power_sims_figures")` was performed. This happens in two steps: (1) create $10,000,000$ observations of $(X,Y,Z,U)$, and (2) summarize the simulated observations to get $P(X|Z)$ and $P(Y|Z)$. 

The first script showed below uses `simulate_data()` to simulate $10,000,000$ observations for one of the combinations of $\beta_X, \gamma_1, \gamma_U$. The other coefficients are set as $\beta_0 = -\beta_X / 2, \beta_U = \gamma_U, \gamma_0 = -\gamma_1$. (See `simulate_data()` for more details on the model.) The results are saved to a file. Note: these files are somewhat large, which is why this script is run for each combination of coefficients separately. This script was run for all 840 combinations. This is very time consuming, so rerun at own risk.

```{r, code = readLines(here::here("scripts/power/power_sims_only.R")), eval = FALSE}
```

The script included below preps the results of the first script by finding the probabilities $P(X|Z)$ and $P(Y|Z)$, and from there the two-sample bounds, and the ATE. Since we have observations on $(X,Y,Z,U)$, the ATE can be calculated. Once again, to limit the amount of memory required, we do this for 10 combinations at a time, and saving the results to files. The script displayed below does this. Again, beware that this is very time consuming. 

```{r, code = readLines(here::here("scripts/power/power_prep_for_plots.R")), eval = FALSE}
```

Finally, we stitch together the 84 .Rds files with all the bounds from the 840 combinations of coefficients into one .Rds file. 

```{r, code = readLines(here::here("scripts/power/power_combine_results.R")), eval = FALSE}
```

